Pragmatics account for limits on theory of mind use in adults

### 2016-01-21 writing/cogsci-submission/cogsci_keysar.tex
edits to intro

\newcommand{\jd}[1]{\green{$^*$}\marginpar{\footnotesize{JD: \green{#1}}}}

@@

Humans can accurately and intelligently reason about the mental states of other humans. Among other things, this ability -- called \emph{theory of mind} \cite{PremackWoodruff78_ChimpanzeeToM} -- allows us to infer the underlying beliefs and intentions that motivate others' actions, and to use these inferences to predict future actions \cite{BakerSaxeTenenbaum09_ActionUnderstandingInversePlanning}. Most children acquire this ability by age six \cite{WimmerPerner83_BeliefsAboutBeliefs, WellmanCrossWatson01_ToMMetaAnalysis} and it serves as an important landmark in the developmental trajectory of intuitive theories \cite{GopnikWellman12_ReconstructingConstructivism}.

While theory of mind use often appears to be automatic and effortless, Keysar and colleagues \cite{KeysarBarr___Brauner00_TakingPerspective, KeysarLinBarr03_LimitsOnTheoryOfMindUse, LinKeysarEpley10_ReflexivelyMindblind} have argued that it is actually the opposite, even for adults: we are ``mindblind" by default and only overcome our egocentric biases through an effortful process of perspective-taking. In other words, while adults are \emph{capable} of applying theory of mind reasoning, we do not always apply it reliably. This argument is based on a simple experimental paradigm, where participants played a simple communication game with a confederate. The two players were placed on opposite sides of a $4 \times 4$ grid containing a set of everyday objects. The confederate played the role of `director,' giving instructions about how to move objects around a grid, and the participant played the role of `matcher,' attempting to follow these instructions. For example, the objects in one trial included a medium-sized measuring cup in one location and a smaller measuring cup in another location. The director gave an instruction like `move the large cup up one square,' referring to the medium-sized one.

Some objects were occluded such that only the matcher could see them, creating an asymmetry in the players' knowledge. To perform accurately, the matcher would need to apply theory of mind and reason about the director?s beliefs. For example, if a much larger measuring cup were placed in an occluded slot and agents failed to account for the director?s beliefs, they might interpret `the large cup' to mean this occluded cup even though the director couldn't possibly know about it. Indeed, \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse} found that participants attempted to move the hidden item in 30\% of cases: 71\% of participants attempted to move this hidden item at least once (out of four critical cases) in the experimental condition, compared to 0\% in a control condition where there was no ambiguity over the referent. 46\% of participants reached for the hidden object at least twice. Additionally, eye-tracking data showed that participants considered the hidden item more often and for longer in the experimental condition than the control condition.

While these results are compelling, the paradigm has been criticized from several different angles. First, in all experiments, the privileged object was a better fit for the referring expression than the one in common ground (e.g. the largest measuring cup vs. the medium measuring cup for ``the large measuring cup"), making the privileged object \emph{a priori} more likely to be the referent. It would be fairer to compare two objects that fit the referring expression equally well \cite{HellerGrodnerTanenhaus08_Perspective}. Second, the viewpoint asymmetry paradigm is somewhat unnatural: common ground is typically built incrementally over the course of an interaction rather than presented all at once, and it is rare for a shared display to differ in perceptual accessibility \cite{HannaTanenhausTrueswell03_CommonGroundPerspective}. Third, none of the experiments reported by Keysar and colleagues included a key comparison condition where the critical item (e.g. the largest measuring cup) was \emph{also} in common ground \cite{BrownSchmidtHanna11_IncrementalPerspectiveTaking}. It would be useful to know whether people referred to the critical item less in the privileged condition than in the full common ground condition, which would put a more positive spin on the results. %Note that none of these criticisms invalidate the paradoxical result that listeners move an occluded object, they just suggest an alternative explanation. Reference disambiguation is probabilistic: common ground and goodness of fit are weighed against one another.

In this paper, we offer a novel pragmatic account of Keysar's results. When both players know that objects are occluded in the display, we note that the speaker is likely to be overinformative. If the listener \emph{expects} the speaker to be overinformative, they will pragmatically pick the \emph{a priori} more likely referent of the referring expression, which in critical trials will be the occluded object. In other words, it is precisely \emph{because} the listener takes the speaker's mental state into consideration that they are tricked by an uncooperative confederate into choosing the wrong item. We began by replicating \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse} (in a multi-player web experiment, recruiting participants to be both director and matcher instead of using a confederate. Instructions for critical items, as well as a random subset of filler items, remained scripted as in the original study. We then ran the same experiment without scripted items, observing unconstrained director behavior. This minimal pair of experiments demonstrates that listener mistakes are at least partially due to the pragmatics of the task, ironically showing that apparent failures of theory of mind are in fact attributable to theory of mind.

>>>>

\newcommand{\jd}[1]{\green{$^*$}\marginpar{\footnotesize{JD: \green{#1}}}}
\definecolor{Green}{RGB}{10,200,100}
\newcommand{\ndg}[1]{\textcolor{Green}{[ndg: #1]}}

@@

Humans can accurately and intelligently reason about the mental states of other humans. Among other things, this ability -- called \emph{theory of mind} \cite{PremackWoodruff78_ChimpanzeeToM} -- allows us to infer the underlying beliefs and intentions that motivate others' actions, and to use these inferences to predict future actions \cite{BakerSaxeTenenbaum09_ActionUnderstandingInversePlanning}. Children acquire this ability by at least age six \cite{WimmerPerner83_BeliefsAboutBeliefs, WellmanCrossWatson01_ToMMetaAnalysis} and it serves as an important landmark in the developmental trajectory of intuitive theory use \cite{GopnikWellman12_ReconstructingConstructivism}.
%
While theory of mind use often appears to be automatic and effortless, Keysar and colleagues \cite{KeysarBarr___Brauner00_TakingPerspective, KeysarLinBarr03_LimitsOnTheoryOfMindUse, LinKeysarEpley10_ReflexivelyMindblind} have argued that it is actually the opposite, even for adults: we are ``mindblind" by default and only overcome our egocentric biases through an effortful process of perspective-taking. In other words, while adults are \emph{capable} of applying theory of mind reasoning, we do not always apply it reliably. 
In this paper we argue that the apparent failures this view is based on are in fact successes for sophisticated social reasoning. 
In particular, we argue that critical utterances used by Keysar and colleagues are \emph{uncooperative}: they are less informative than what a speaker would actually produce in that situation; listeners expect these more informative utterances, and interpret the actual utterances accordingly---resulting in ``errors''.

The argument of Keysar and colleagues is based on an elegant experimental paradigm, where participants played a simple communication game with a confederate. The two players were placed on opposite sides of a $4 \times 4$ grid containing a set of everyday objects.
\ndg{it'd be nice to have a grid illustrating one of the conditions to refer to here...}
The confederate played the role of `director,' giving instructions about how to move objects around a grid, and the participant played the role of `matcher,' attempting to follow these instructions. For example, the objects in one trial included a medium-sized measuring cup in one location and a smaller measuring cup in another location. The director gave an instruction like `move the large cup up one square,' referring to the medium-sized one. 
Some objects were occluded such that only the matcher could see them, creating an asymmetry in the players' knowledge. To perform accurately on critical trials, the matcher would need to apply theory of mind to reason about the director's beliefs. For example, imagine a much larger measuring cup were placed in an occluded slot: if a participant failed to account for the director's (partial) knowledge, she might interpret `the large cup' to mean this occluded cup (which the director couldn't possibly know about). 
Indeed, \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse} found that participants attempted to move the hidden item in 30\% of cases: 71\% of participants attempted to move this hidden item at least once (out of four critical cases) in the experimental condition, compared to 0\% in a control condition where there was no ambiguity over the referent. 
%46\% of participants reached for the hidden object at least twice. 
Additionally, eye-tracking data showed that participants considered the hidden item more often and for longer in the experimental condition than the control condition.

While these results are compelling, the paradigm has been criticized from several different angles. First, in all experiments, the privileged object was a better fit for the referring expression than the one in common ground (e.g. the largest measuring cup vs. the medium measuring cup for ``the large measuring cup"), making the privileged object \emph{a priori} more likely to be the referent. It would be fairer to compare two objects that fit the referring expression equally well \cite{HellerGrodnerTanenhaus08_Perspective}. Second, the viewpoint asymmetry paradigm is somewhat unnatural: common ground is typically built incrementally over the course of an interaction rather than presented all at once, and it is rare for a shared display to differ in perceptual accessibility \cite{HannaTanenhausTrueswell03_CommonGroundPerspective}. Third, none of the experiments reported by Keysar and colleagues included a key comparison condition where the critical item (e.g. the largest measuring cup) was \emph{also} in common ground \cite{BrownSchmidtHanna11_IncrementalPerspectiveTaking}---perhaps people would refer to the critical item less in the privileged condition than in the full common ground condition, which would put a more positive spin on the results. %Note that none of these criticisms invalidate the paradoxical result that listeners move an occluded object, they just suggest an alternative explanation. Reference disambiguation is probabilistic: common ground and goodness of fit are weighed against one another.

In this paper, we offer an additional factor that helps account for Keysar's results. 
Theory of mind as applied within language understanding depends on an accurate model of what a speaker would say in different situations; given an utterance, a listener can then reason backward to the most plausible situation.
This suggests that we consider whether the utterances used by the confederate in Keysar's critical conditions were actually what a speaker would be expected to say. 
If not, then perhaps participants are making choices that are in fact consistent with correct pragmatic interpretation on the confederate's (uncooperative) utterance.
More precisely, when both players know that objects are occluded in the display, the speaker may tend to add additional precision to references in order to avoid confusion. If the listener \emph{expects} the speaker to do this, they will pragmatically pick the \emph{a priori} more likely referent of the referring expression, which in critical trials will be the occluded object. In other words, it is precisely \emph{because} the listener takes the speaker's mental state into consideration that they are tricked by an uncooperative confederate into choosing the wrong item.

We began by replicating \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse} in a multi-player web experiment. We recruited participants to be both director and matcher (instead of using a confederate), but instructions for critical items, as well as a random subset of filler items, remained scripted as in the original study. We replicated the original finding, but noted a tendency of directors to be ove-rinformative in unscripted trials. We then ran the same experiment without scripting the directions for critical trials, observing unconstrained director behavior. We observe much greater precision in unconstrained director utterances and better performance of the matchers. This minimal pair of experiments demonstrates that listener mistakes are at least partially due to the pragmatics of the task, ironically showing that apparent failures of theory of mind are in fact attributable to sophisticated expectatuions about speaker behavior---that is, to theory of mind.

\ndg{intro can probably be compressed some if needed.}
